Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{DISTRIBUTION STATEMENT A. Approved for public release.}

### 2023-02-14 paper/paper.tex
Inline the high assurance contracts sections

The AGREE specification for the SW component in \secref{sec:example} is given in \figref{fig:sw}.
%
\linesref{line:sw-event-def-start}{line:sw-event-def-end} define local variables that are true when data is present on the corresponding event ports for the component.  These variables are
purely for notational convenience.
%
The contract has three assumptions:
%
\begin{itemize}
\item \lineref{line:sw-assume-1} and \lineref{line:sw-assume-2}.
When input is present it is well-formed. The well-formed predicates
are defined elsewhere using AGREE functions.
\item \lineref{line:sw-assume-3}. If there is an incoming request,
it is either the very first one, \texttt{\textbf{Historically}(\textbf{not} req)},
or it arrives after the component has output a start event in response to a previous
request, \texttt{\textbf{Since}(\textbf{not} req, strt)}---to be read as \emph{not
request since start}.
\end{itemize}

@@

\section{Code contracts for high assurance components}
\label{sec:code-contracts}
\input{high-assurance-contracts}

>>>>

The AGREE specification for the SW component in \secref{sec:example} is given in \figref{fig:sw}.
%
\linesref{line:sw-event-def-start}{line:sw-event-def-end} define local variables that are true when data is present on the corresponding event ports for the component.  These variables are
purely for notational convenience.
%
The contract has three assumptions:
%
\begin{itemize}
\item
\lineref{line:sw-assume-1} and \lineref{line:sw-assume-2}.
%
When input is present it is well-formed.
%
The well-formed predicates are defined elsewhere using AGREE functions.

\item
\lineref{line:sw-assume-3}.
%
If there is an incoming request, it is either the very first one, \texttt{\textbf{Historically}(\textbf{not} req)}, or it arrives after the component has output a start event in response to a previous request, \texttt{\textbf{Since}(\textbf{not} req, strt)}---to be read as \emph{not
request since start}.
\end{itemize}

@@

\section{Code contracts for high assurance components}
\label{sec:code-contracts}

Generally, AGREE contracts do not describe the computation that a component performs.
%
This is entirely by design: AGREE is intended to be declarative and reasoning about component behavior takes place solely at the specification level.
%
It defines what a component computes and not how it computes it.
%
However, the syntax of AGREE provides enough expressiveness to support the notion of a \emph{code contract}: a contract from which an implementation can be extracted.
%
First we must discuss a class of guarantees, \emph{output guarantees}, which determine the values on all output ports of a component.

\begin{definition}[Output guarantee]
An \emph{output guarantee} is a stylized guarantee that fully specifies the data written to an output port.
%
There are three possibilities according to whether the output port $p$ is a \konst{data} port, an \konst{event} port, or an \konst{event data} port:
%
\[
\begin{array}{ll}
\konst{data}: &  p = \mathit{e} \\
\konst{event}: &  \konst{event} (p) = \mathit{b} \\
\konst{event data}: & \itelse{b}{\konst{event} (p) \land p = e}{\neg \konst{event}(p)} \\
\end{array}
\]
%
\end{definition}

Informally, a code contract treats its \konst{eq} ``statements'' as defining a list of assignments to state variables, and its output guarantees as directives for producing output.

\begin{definition}[Code contract]
A leaf component of the form $(I,O,A,P,\emptyset)$ is a \emph{code contract} if $\mathit{Eqs} \cup G \subseteq P$, where \[\mathit{Eqs} = \set{v_1 = e_1, \cdots , v_n = e_n} \] is a non-empty set of \konst{eq} statements and $G$ is the set of output guarantees, one for each element of $O$.
%
In the interpretation as code, the order of elements of $\mathit{Eqs}$ is important, and is simply taken to be the occurrence order of the \konst{eq} statements in the syntax.
%
Thus we will work with the \emph{list} of equations $\mathit{Eqs} = [v_1 = e_1; \cdots ; v_n = e_n]$.
\end{definition}

The semantics of \secref{sec:agree-semantics} supports a formal connection between the original contract, and verification results of the AGREE model checker being run on it, and code generated from specifications of leaf level cyber-components.
%
It also provides the root meaning at the base of a chain of translation steps moving from an AGREE contract to a CakeML executable.

The first step in the chain maps the contract to a code-focused representation.
Assume given a code contract $(I,O,A,\mathit{Eqs} \cup G \cup P,\emptyset)$, environment $E$, and time $t$.
%
The \emph{evaluation} of $\mathit{Eqs}$ followed by the evaluation of $G$ results in a new environment $E'$ where the stream values for state variables and outputs have been computed for time $t$.
%
The step from $E$ to $E'$ is one full cycle in the repeated evaluation of the component.

\begin{definition}[Evaluation]
We overload existing notation and write the evaluation of $\mathit{Eqs}$ and then $G$ in environment $E$ at time $t$ as $\sem{\mathit{Eqs}\cdot G}^E_t$.
%
The evaluation of $v = e \in {\Eqs}$ is an environment transformation
%
\[
 \sem{v = e}^E_t = E[v(t) \mapsto\sem{e}^E_t] \ .
\]
%
modifying $E$ so that stream $v$ has value $\sem{e}^E_t$ at time $t$.
%
The list {\Eqs} is evaluated by folding the transformation left-to-right through it.
%
The transformation is similarly applied to the list of output guarantees $G$, computing the values on output ports.
%
The details are omitted, being a bit messy because of the different kinds of output ports available.
\end{definition}

\begin{definition}[Code contract correctness]
Assume code contract $(I,O,A,\mathit{Eqs} \cup G \cup P,\emptyset)$.
%
The contract is \emph{correct}, if for all $E$ and $t$, whenever the assumptions (historically) hold in $E$ and evaluation steps from $E$ to $E'$, then the guarantees hold in $E'$:
%
\[
\sem{\konst{Hist}(A)}^E_t \land E' = \sem{\mathit{Eqs} \cdot G}^E_t \imp \sem{P}^{E'}_t
\]
%
\end{definition}

How does this notion of correctness integrate with the verification conditions generated by AGREE?
%
In \emph{contract correctness}, the model checker is proving the following property
%
\[
\konst{Hist}(A) \imp P
\]
%
and in \emph{code contract correctness} we essentially prove a Hoare triple of the form
%
\[
\set{\konst{Hist}(A)}\; (\mathit{Eqs} \cdot G) \; \set{P}
\]
%
This accords with intuition, namely that AGREE is a kind of \emph{program-free program logic};
%
by pulling out a program $(\mathit{Eqs}\cdot G)$ from a code contract we find ourselves in the domain of imperative program verification.
%
For example, at this point one could apply a verification condition generator to help break down the proof obligation.
%
In order to use the code contract result in system contract verification conditions, one has to show that the code contract result implies the system verification condition.

%%
% IMPERATIVE CODE CONTRACTS
%%

\subsubsection*{Imperative code contracts}

Stream-based evaluation allows values arbitrarily \emph{deep in the past} to be accessed by use of $\konst{pre}(-)$.
%
However, the programming languages targeted by SPLAT do not directly support streams.
%
We are developing a translation to eliminate such deep temporal accesses by the introduction of intermediate variables.
%
The essential property the translation needs to satisfy is a syntactic restriction, call it \emph{Pascal-like}, that supports easy translation to standard imperative languages.
%
The guiding intuition is that a Pascal-like {\Eqs} accesses at most one step in the past;
%
also, accesses to a past value of a variable are only possible until the variable is updated.
%
In informal terms, this means that ${\Eqs}\cdot G$ in a code contract has the following properties:
%
\begin{itemize}
\item Followed-by only occurs at the top level of a variable definition, \eg, $v = e_1 \to e2$;
\item $\konst{pre}(e)$ is allowed only when $e$ is a variable; and
\item $\konst{pre}(v)$ is allowed only before the defining equation $v = e$ (and may occur in $e$).
\end{itemize}
%
For example, the Fibonacci sequence $1,1,2,3,5,8,13,\ldots$ can be defined, albeit in a manner that is not Pascal-like, by
%
{\small
\begin{lstlisting}[style=agree]
  Fib = 1 -> pre(1 -> Fib + pre Fib)
\end{lstlisting}
}
%
\noindent or by (again this is not Pascal-like)
%
{\small
\begin{lstlisting}[style=agree]
  N = 0 -> 1 + pre(N)
  Fib = if N <= 1 then 1 else pre(Fib) + pre(pre(Fib))
\end{lstlisting}
}
%
\noindent or by (this time Pascal-like):
%
{\small
\begin{lstlisting}[style=agree]
  N = 0 -> 1 + pre(N)
  F2 = 42 -> pre(F1)
  F1 = 42 -> pre(Fib)
  Fib = if N <= 1 then 1 else F1 + F2
\end{lstlisting}
}
%
Please note that the initial values of \verb+F1+ and \verb+F2+ are never accessed, so 42 could be replaced by any number in their defining equations.
%
A Pascal-like ${\Eqs} \cdot G$ is straightforward to implement in conventional programming languages, as we will see in \secref{sec:synthesis}.

\subsection{Code contracts for the hardened system}
\label{sec:ha-contracts}

\newsavebox{\flt}
\begin{lrbox}{\flt}
  \begin{lstlisting}[style=agree,numbers=left] -- start user
    -- start user definitions
    eq policy : bool =
      WELL_FORMED_AUTOMATION_RESPONSE(Input);
    -- end user definitions

    guarantee "Filter output is well-formed" :
      if event(Input) and policy then
        event(Output) and Output = Input
      else
        not event(Output);
  \end{lstlisting}
\end{lrbox}

\newsavebox{\mntr}
\begin{lrbox}{\mntr}
  \begin{lstlisting}[style=agree,numbers=left]
    const is_latched : bool = true;

    -- start user definitions
    assume "One automation request in flight at a time" : *\label{line:mon-assume}*
      true -> (req => pre(Historically(not req) or Since(not req, rsp)));

    const MAX_LATENCY : int = 1; *\label{line:mon-const}*

    eq rsp : bool = event(Response);
    eq req : bool = event(Request);

    eq preIsPending : bool = false -> pre(isPending);*\label{line:mon-pre-pending}*
    eq isPending : bool = Since(not rsp, req and not rsp);*\label{line:mon-pending}*
    eq latency : int = 0 -> (if req then 0 else pre(latency) + 1);*\label{line:mon-latency}*

    eq policy : bool = (rsp => req) -> *\label{line:mon-policy}*
                         (    (isPending => latency < MAX_LATENCY)
                          and (rsp => (req or preIsPending)));
    -- end user definitions

    eq alert : bool = (not policy) -> *\label{line:mon-alert}*
                        ((is_latched and pre(alert)) or not policy);

    guarantee "Alert port tracks alert variable" :
      event(Alert) = alert;
    guarantee "Output if not alerted" :
      if (not(alert) and rsp) then
          event(Output) and (Output = Response)
      else
          not (event(Output));
  \end{lstlisting}
\end{lrbox}

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \scalebox{0.62}{\usebox{\flt}} \\
    \end{tabular}
  \end{center}
  \caption{Code contract for high-assurance filter.}
  \label{fig:filter}
\end{figure}

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
    \scalebox{0.62}{\usebox{\mntr}} \\
    \end{tabular}
  \end{center}
  \caption{Code contract for high-assurance monitor.}
  \label{fig:monitor}
\end{figure}

Code contracts for the filter and monitor are shown in \figref{fig:filter} and \figref{fig:monitor}.
%
The filter is direct: it defines the policy according to the well-formed requirement for data.
%
We now discuss the monitor in some detail.

The monitor informally behaves as follows: if the \emph{policy} is obeyed in the current step, and the alert has not yet been raised, then the \texttt{Response} input is sent out on \texttt{Output}, and the \texttt{Alert} output stays low \lineref{line:mon-alert}.
%
Otherwise \texttt{Alert} goes high and stays high, and there is nothing sent henceforth on \texttt{Output}.

The \texttt{alert} variable (distinct from the \texttt{Alert} port) depends not just on the policy but also on the \texttt{is\_latched} constant, which has been set at configuration time.
%
Since \texttt{is\_latched} is \konst{true}, the alert is persistent, meaning that once the alert is raised, it is always raised;
%
otherwise, it is the complement of the policy value in the current step.

The monitor policy is defined by marking, with variable \texttt{isPending}, when a request is made but not yet satisfied, and by counting the number of steps (\texttt{latency}) between requests.
%
See \linesref{line:mon-pre-pending}{line:mon-latency}:
%
The \texttt{isPending} variable is true whenever there is a request that does not coincide with a response, \ie, \texttt{isPending} holds when \emph{there is no response since the most recent request without a corresponding response}.
%
The \texttt{latency} variable starts at zero, resets on every request, and otherwise increments by one.  Note that the \texttt{preIsPending} variable is a temporary needed to make the contract Pascal-like;
%
otherwise, the current and previous value of \texttt{isPending} would be simulataneously accessed in the definition of \texttt{policy}.

The policy (\lineref{line:mon-policy}) makes sure the latency is bounded whenever there is a pending request, and it ensures that a response is a result of a request.
%
The initial step differentiates from later steps in only needing to be sure a response is a result of a request, both happening in the initial step.

The latency bound for the monitor is defined by the constant \texttt{MAX\_LATENCY} (\lineref{line:mon-const}).
%
In this example the constant is set to one to be consistent with the system specification from the previous section.
%
At time zero the policy is trivial: a response requires an accompanying request.
%
After time zero, if a request is pending in the current time step, then the latency must still be under the bound, and if there is a response, then it closes a request now or a pending request from earlier in time.

The policy definition only works if there is never more than one outstanding request at a time;
%
otherwise, the latency counter resets incorrectly.
%
That requirement is encoded in the assumption (\lineref{line:mon-assume}) and is the same assumption used for the system input.
%
In fact, the policy can be written without the assumption, but the policy is much simpler to write with the assumption.

As shown in \figref{fig:hardened-certificate}, the cyber-hardened system passes AGREE verification.
%
That means that the high-assurance components protect the WM from the untrusted AI generating malformed data, spontaneous responses, or delayed responses.
%
Synthesis must generate faithful implementations of the code contracts for the AGREE verification result to apply to the deployed system.
